A Fast and Tight Heuristic for A* in Road Networks

### 2021-08-09 paper/ch_potentials.tex
reordering, some a* and preliminary writing

\section{Algorithm}\label{sec:main-algo}

@@

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G+$ (blue) exists.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

Using the CH query algorithm, we can already give a simple heuristic.
The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
This yields tight estimates but a high overhead for the heuristic evaluation.
While a single CH query is fast, answering one for every node explored in the $A^*$ search is slow.
Fortunately, we can do better.
% For this we need another component called PHAST~\cite{dgnw-phast-13}.

\subsection{PHAST based Heuristic}

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
The preprocessing phase remains unchanged.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the shortest distances from all nodes to $t$ were computed.
%PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
%PHAST's main advantage is that can easily be parallelized.
%However, we will not consider parallelization in this paper.
%We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
Using PHAST, we can also compute a tight A* heuristic.
In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

% This PHAST-based algorithm works.
The $H$ lookup and by extension the $A^*$ search is indeed fast.
However, the PHAST step before the search is comparatively expensive.
The reason is that the distances towards $t$ are computed for \emph{all} nodes.
Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.

\subsection{CH-Potentials}

Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
In a first step, we run the backward CH search from $t$ to obtain an array $B$.
$B[x]$ is the minimum down $xt$-path distance or $+\infty$, if there is no such path.
$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.

To compute the heuristic $h_t(x)$, we recursively compute for all up-edges $(x,y)$ the heuristic $h_t(y)$.
Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_y\{w_\ell(x,y) + h_t(y)\}$.
As not all shortest up-down paths contain an up-edge, we set $h_t(x) = \min \{ B[x], d \}$.
This calculation is correct, as it computes the minimum up-down $xt$-path distance in $G^+_\ell$, which corresponds to the minimum $xt$-path distance in a CH.
A* with this heuristic is the basic CH-Potentials algorithm.

\subsection{CCH-Potentials}

\section{Low Degree A* Improvements}

\label{sec:low-deg-improvment}

@@

\subsection{Skip Degree Two Nodes}

@@

\subsection{Skip Degree Three Nodes}

@@

\subsection{Stay in Largest Biconnected Component}
\label{sec:largested-biconnected-component}

@@

\section{Applications}
\label{sec:extensions}

@@

\subsection{Avoiding Tunnels and/or Highways}
\label{sec:no-tunnel-highway}

@@

\subsection{Forbidden Turns and Turn Costs}
\label{sec:no-turns}

@@

\subsection{Predicted Traffic or Time-Dependent Routing}
\label{sec:predicted-traffic}

@@

\subsection{Live and Predicted Traffic}
\label{sec:live-predicted-traffic}

@@

\subsubsection{Three-Phase Setups}

@@

\subsection{Temporary Driving Bans}

@@

>>>>

\section{Preliminaries}\label{sec:preliminaries}

We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
Given nodes $s$ and $t$ we want to obtain the path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
We denote this shortest path weight as the distance $\operatorname{dist}_w(s,t)$ between $s$ and $t$.

\subsection{Dijkstra's Algorithm}

Dijkstra's algorithm can be used to solve this point to point shortest path problem.
It maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
Initially, all distances are set to $\infty$.
The start node distance $d[s]$ is set to zero and the queue initialized with $s$.
Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
Outgoing edges $uv$ are relaxed, i.e. $d[v] = \min(d[v], d[u] + w(uv))$.
If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
Once $t$ was settled, $d[t]$ contains the distance between $s$ and $t$.

\subsection{A*}

A* is a goal-directed variant of Dijkstra's algorithm.
% A* uses a \emph{heuristic} to guide the search towards the goal.
It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target.
The priority queue is then ordered by $d[v] + h_t(v)$.
Thus, nodes closer to the target a explored earlier.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A*'s running time crucially depends on how tight this estimate is.

A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights} $w_{h_t}$.
This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
A heuristic function is \emph{feasible} iff $w_{h_t} \geq 0$.

A CH places nodes into levels.
No edge must connect two nodes within one level.
Levels are ordered by ``importance''.
The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
An edge goes \emph{up} when it goes from a node in a lower level to higher level.
\emph{Down} edges are defined analogously.
An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
$m$ is called the \emph{mid} node.
An \emph{up path} is a path where the last node is the mid node.
Similarly, the first node is the mid node of a \emph{down path}.
% Every up and down path is an up-down path.
%
In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
%See~\cite{gssv-erlrn-12} for the details.
After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
See Figure~\ref{fig:ch} for a proof sketch.
From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) exists.
Thus, we can restrict our search to up-down paths in $G^+$.
The search is bidirectional.
The forward search starts from $s$ and only follows up-edges.
Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
The two searches meet at the mid node.
Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
The forward search works analogously.
%
A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.

% Using the CH query algorithm, we can already give a simple heuristic.
% The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
% This yields tight estimates but a high overhead for the heuristic evaluation.
% While a single CH query is fast, answering one for every node explored in the $A^*$ search is slow.
% Fortunately, we can do better.
% For this we need another component called PHAST~\cite{dgnw-phast-13}.

\subsection{Customizable Contraction Hierarchy (CCH)}

% need elimination tree

\subsection{PHAST}

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the one-to-all case works analogously).
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable nodes via reversed down-edges are explored.
Algorithm~\ref{algo:ch-backward} shows this first step.
The second step iterates over all CH levels from top to bottom.
In each iteration, all up-edges starting within the current level are followed in reverse.
After all levels are processed, the shortest distances from all nodes to $t$ were computed.
Pseudo-code is provided in Algorithm~\ref{algo:phast}.
PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
It can also easily be parallelized.
% However, we will not consider parallelization in this paper.
We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
% Using PHAST, we can also compute a tight A* heuristic.
% In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
% Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.

% This PHAST-based algorithm works.
% The $H$ lookup and by extension the $A^*$ search is indeed fast.
% However, the PHAST step before the search is comparatively expensive.
% The reason is that the distances towards $t$ are computed for \emph{all} nodes.
% Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.

\subsection{RPHAST}

RPHAST, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node.
Given a set of source nodes $S$ and a target node $t$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
This step is called \emph{selection}.
Then, the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the new restricted subgraph.
RPHAST is particularly effective when many targets are queried for the same source set $S$.
If the source set changes often, selection times may become problematic.

\section{The Incremental Many-to-One Problem}\label{sec:lazy-rphast}

% TODO not so great intro directly after PHAST/RPHAST
Route planning algorithms typically focus on the point-to-point shortest path problem.
However, certain applications need distances between sets $S$ and $T$ of nodes.
This is variant is known as the \emph{many-to-many} problem variant.
In some cases on of the sets is a single node.
This is then called a \emph{one-to-many} or \emph{many-to-one} problem.
In this section, we study (discuss) a very specific setting arises naturally from A* heuristics.
We denote this problem as the \emph{incremental many-to-one} problem.
The first step is the \emph{target selection} where the target node is provided.
Then, an arbitrary number of source nodes $s_i$ are given one after another and for each source the distance to the target has to be computed.

We consider the combined running time of the target selection and each incremental query as the running time to answer such an incremental many-to-one query.
Since the target selection time is included, computing the distances to all nodes with Dijkstra or PHAST is too slow.
Fortunately, we can do better.

\subsection{Lazy RPHAST}

The core idea of our algorithm is to do the PHAST computation lazily and using memoization as depicted in Algorithm~\ref{algo:pot}.
In the target selection, we first run the backward CH search from $t$ to obtain an array $B$.
$B[x]$ is the minimum down $xt$-path distance or $+\infty$, if there is no such path.
$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.

To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i,x)$ the distance $D[x]$.
Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i,x) + D[x]\}$.
As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
% TODO consistent uv vs u,v
% A* with this heuristic is the basic CH-Potentials algorithm.

\subsection{Lazy RPHAST on CCH}

\section{A* with CH-Potentials}\label{sec:main-algo}

@@

\subsection{CH-Potentials Heuristic}

CH-Pot is A* with Lazy RPHAST Pot.
Given $s$ and $t$ upward search from $t$
Then compute $h_t(x)$ for with ComputeAndMemoizeDist

\subsection{Bidirectional A*}

Bidir A* not as straightforward which one might expect.
As pointed out in~\cite{gh-cspas-05}, one can not simply use a forward potential $h_t$ for the forward search and a backward potential $h_s$ for the reverse search.
Since these searches would operate on different reduced graphs, it is unclear when the searches can terminate.
To the best of our knowledge, no better stopping criterion exists than running both searches until the minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
That is the same stopping crit as for a unidir search..
symmetric
Thus, the symmetric bidirectional algorithm performs double the work than a simple unidir search.
can be slightly improved through pruning $\overrightarrow{D[x]} + h_t(x) > \mu$.
Still not competitive

average potential - reduced graph with average of reduced edge weights
but average pot in each dir worse than orig pot for dir
also overhead

we propose an improved pruning crit, one which takes the search progress into account (through the min queue keys), which makes the simpler symmetric approach competitive or even faster than the average pot
If $x$ was not yet settled by reverse search with minimum queue key $\overleftarrow{k}$ then
$\operatorname{dist}(x,t) \geq \overleftarrow{k} - h_s(x)$.
Thus, if $\overrightarrow{D[x]} + \overleftarrow{k} - h_s(x) > \mu$ we can discard $x$.

% direction switch strats!

\subsection{Low Degree A* Improvements}\label{sec:low-deg-improvment}

@@

\subsubsection{Skip Degree Two Nodes}

@@

\subsubsection{Skip Degree Three Nodes}

@@

\subsubsection{Stay in Largest Biconnected Component}\label{sec:largested-biconnected-component}

@@

\subsection{Applications}\label{sec:extensions}

@@

\subsubsection{Avoiding Tunnels and/or Highways}
\label{sec:no-tunnel-highway}

@@

\subsubsection{Forbidden Turns and Turn Costs}
\label{sec:no-turns}

@@

\subsubsection{Predicted Traffic or Time-Dependent Routing}
\label{sec:predicted-traffic}

@@

\subsubsection{Live and Predicted Traffic}
\label{sec:live-predicted-traffic}

@@

\subsubsection{CCH Potentials}
% TODO CCH-Pot

@@

\subsubsection{Temporary Driving Bans}

@@

\subsubsection{Alternative Routes with the Penalty Method}

% TODO
